Phylogenetics and the human microbiome.

### 2013-10-03 sbreview.tex
Looking pretty good.

Several groups are continually revising taxonomy using the results of phylogenetic tree inference.
These attempts are less ambitious than the development of the PhyloCode \citep{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while for the most part leaving taxonomic names fixed.
% to preserve information associated with taxonomic classifications.
Bergey's Manual of Systematic Bacteriology has officially adopted 16S as the basis for their taxonomy, although the actual revision process appears opaque \citep{kreig1984bergey}.
The GreenGenes group \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved}.
Tax2tree uses a heuristic algorithm to optimize the $F$-measure of precision and recall for its taxonomic assignments.
Interestingly, tax2tree allows for polyphyletic taxonomic groups as allowance for either phylogenetic error, lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis; the method does not attempt to signal the cause of such polyphyletic groups.
\citet{matsen2012reconciling} developed algorithms to quantify discordance between phylogeny and taxonomy based on a coloring problem previously described in the computer science literature \citep{moran2008convex}.
Although it is wonderful that several groups are actively working on taxonomic revision, it can be frustrating to have multiple different taxonomies with no easy way to translate between them or to the taxonomic names provided in the NCBI sequence database.

An obvious application of phylogenetics is to perform taxonomic classification, as the taxonomy is at least in part defined by phylogeny.
However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple classifiers based on $k$-mer composition \citep{wang2007naive,rosen2008metagenome}.
\citet{srinivasan2012bacterial} find that phylogenetic methods to do taxonomic classification can outperform composition-based techniques at least for certain taxonomic groups.
% \citep{werner2011impact} Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
The MEGAN program \citep{huson2007megan,huson2011integrative} uses a least common ancestor (LCA) strategy on taxonomies to derive taxonomic assignments for that sequence from BLAST results of an unknown sequence onto a database of sequences with taxonomic labels.
\citet{munch2008statistical,munch2008fast} infer taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence.
\citet{segata2012metagenomic} propose a clever approach to inferring organisms present in a metagenomic sample by compiling a database of clade-specific genes, then classifying a given read as being from the only clade that has the corresponding gene.
They show that this has good sensitivity and specificity, however, this method can only be used to identify the presence of organisms whose genome has been sequenced.

@@

@@

\subsection{Diversity estimates using phylogenetics}
Because 16S surveys are inherently complex and noisy data, summary statistics are often used; summaries of the diversity of a single sample are often called \emph{alpha diversity}.
For the most part, this literature adapts methods from the classical ecological literature by substituting OTUs for taxonomic groups.
%The most common diversity indices are Simpson and Shannon.
% \citep{hill1973diversity} Diversity and evenness: a unifying notation and its consequences
However, phylogenetic diversity metrics are also used and here we will focus on their applications and methods.

Despite the enthusiasm with which microbial ecologists have accepted between-community comparison using phylogenetics (next section), phylogenetic alpha diversity seems under-developed.
Whereas just about every 16S survey investigation involves an OTU-based alpha diversity estimate, only a few involve phylogenetic diversity (PD) measures, and only in the sense of \citet{faith1992conservation}.
Faith's PD has been applied to some 16S survey data \citep{lozupone2007global,costello2009bacterial} and to metagenomic reads in a set of marker genes \citep{kembel2011phylogenetic}.

Although abundance-weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are among the most common, abundance-weighted phylogenetic diversity measures are not used in human microbiome studies.
Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
\citet{mccoy2013abundance} have recently shown that partially abundance-weighted diversity measures do a good job of distinguishing between dysbiotic and "normal" states of the human microbiome; in particular that they do a better job than the commonly-used OTU-based measures.
\citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths.

@@

\newcommand{\unifracLegend}{
    The UniFrac divergence measure \citep[figure adapted from][]{LozuponeKnightUniFrac05}.
    When samples are interspersed across the tree (left tree), they have a smaller \textit{fraction} of branch lengths that can be \textit{uniquely} assigned to one sample or another compared to when they are separate (right tree).
}
\forarxiv{
\begin{wrapfigure}[23]{r}{2.5in}
  % \capstart
  \vspace{-19pt}
  \begin{center}
    \includegraphics[width=2.5in]{figures/unifrac.pdf}
  \end{center}
  \vspace{-10pt}
  \caption{\unifracLegend}
  \label{fig:unifrac}
\end{wrapfigure}
}

@@

Alignment methods are primarily focused on extending a curated ``seed alignment'' by adding additional sequences; several tools have been created with exactly this application for 16S in mind \citep{desantis2006nast,caporaso2010pynast,pruesse2012sina}.
The community also uses profile hidden Markov models \citep{eddy1998profile} and CM models \citep{nawrocki2009infernal,nawrocki2009structural} to achieve the same result.

The large data sets associated with human microbiome analysis require highly efficient algorithms for \emph{de novo} tree inference.
Historically this has meant relaxed neighbor joining \citep{evans2006relaxed}, but more recently FastTree 2 \citep{price2010fasttree} has emerged as the \textit{de facto} standard.
People do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has ported in the clearcut \citep{sheneman2006clearcut} program, or QIIME \citep{caporaso2010qiime}, which wraps clearcut and FastTree.
% AXIOME streamlines and manages analysis of small subunit (SSU) rRNA marker data in QIIME and mothur \citep{lynch2013axiome}.
% Bayesian phylogenetic inference is absent.

@@

Considerable effort goes to the creation of large curated alignments and phylogenetic trees on 16S.
There are two major projects to do so: one is the SILVA database \citep{pruesse2007silva,quast2013silva}, and the other is the GreenGenes database \citep{desantis2006greengenes,mcdonald2011improved}.
Because of the high rate of insertion and deletion of nucleotides in 16S, these alignments have a high percentage of gap.
Taking the length of 16S to be 1543 nt,
% SILVA
% 1543 / 42284
% .96353230536373096207
the 479,726 sequence SILVA reference alignment version 115 is over 96\% gap,
% GreenGenes
% 1543 / 7682
% .79914084873730799272
while the 1,262,986 sequence GreenGenes 13\_5 alignment has is almost 80\% gap.
The SILVA-associated 'all-species living tree' project \citep{yarza2008all} started with a tree inferred by maximum likelihood and has been continually updated  by inserting sequences via parsimony.
The GreenGenes tree is updated by running FastTree from scratch for every release.
There appears to be a commonly held belief that FastTree in particular works well even with such gappy alignments \citep[e.g.][]{sharpton2011phylotu}.

@@

It appears that neutral models involving phylogenetics could be more fully developed.
Methods explicitly invoking trait evolution are notably absent, with the recent exception of \citep{langille2013predictive}.
The results of this simple method are reasonable, but would a species tree reconciled with a collection of gene trees allow for better prediction?
Perhaps improved methods, say involving whole-genome evolutionary modeling or models of metabolic network evolution, could shed light on the problem.
Here again the ``tree of life'' problem can be formulated in a practical light: what representation allows for the best prediction of features of underlying genomes?
How to formulate a useful notion of independent contrasts \citep{felsenstein1985phylogenies} on such an object?
It is quite possible that inference using a more complete representation would not be able to overcome the inherent noise of the data, but further exploration seems warranted as simple methods give reasonable results.

@@

\section{Acknowledgements}
I thank Olivier Gascuel for the opportunity to present on this subject during the 2013 LIRMM Mathematical and Computational Evolutionary Biology workshop and for organizing a corresponding special section of \textit{Systematic Biology}.
I am grateful to Aaron Darling, David Fredricks, Noah Hoffman, Steven Kembel, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, thank to Noah Hoffman, Christopher Small and Bj\"orn Winckler for providing feedback on the manuscript, and thank the NIH (R01 HG005966-01) and NSF (1223057) for financial support.

\notforarxiv{
\newpage
\section{Figure Legends}
%\FIGmassTransport

\noindent
Figure 1: \unifracLegend

\noindent
Figure 2: \dirtpilesLegend

>>>>

Several groups are continually revising taxonomy using the results of phylogenetic tree inference.
These attempts are less ambitious than the PhyloCode project \citep[to develop a taxonomic scheme expressed directly in terms of a phylogeny; see][]{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while for the most part leaving taxonomic names fixed.
% to preserve information associated with taxonomic classifications.
Bergey's Manual of Systematic Bacteriology has officially adopted 16S as the basis for their taxonomy, although the actual revision process appears opaque \citep{kreig1984bergey}.
The GreenGenes group \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved}.
Tax2tree uses a heuristic algorithm to optimize the $F$-measure of precision and recall for its taxonomic assignments.
Interestingly, tax2tree allows for polyphyletic taxonomic groups as allowance for either phylogenetic error, lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis; the method does not attempt to signal the cause of such polyphyletic groups.
\citet{matsen2012reconciling} developed algorithms to quantify discordance between phylogeny and taxonomy based on a coloring problem previously described in the computer science literature \citep{moran2008convex}.
Although it is wonderful that several groups are actively working on taxonomic revision, it can be frustrating to have multiple different taxonomies with no easy way to translate between them or to the taxonomic names provided in the NCBI sequence database.

An obvious application of phylogenetics is to perform taxonomic classification, as the taxonomy is at least in part defined by phylogeny.
However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple classifiers based on $k$-mer composition \citep{wang2007naive,rosen2008metagenome}.
\citet{srinivasan2012bacterial} find that phylogenetic methods to do taxonomic classification can outperform composition-based techniques at least for certain taxonomic groups.
% \citep{werner2011impact} Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
The MEGAN program \citep{huson2007megan,huson2011integrative} BLASTs an unknown sequence onto a database of sequences with taxonomic labels and assigns the sequence the lowest (i.e.\ narrowest) taxonomic group shared by all of the high-quality hits.
\citet{munch2008statistical,munch2008fast} infer taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence.
\citet{segata2012metagenomic} propose a clever approach to inferring organisms present in a metagenomic sample by compiling a database of clade-specific genes, then classifying a given read as being from the only clade that has the corresponding gene.
They show that this has good sensitivity and specificity, however, this method can only be used to identify the presence of organisms whose genome has been sequenced.

@@

\newcommand{\pdLegend}{
    Unweighted phylogenetic diversity (PD, left) and an abundance-weighted PD measure (right), where taxa present in a sample are shown as blue circles and abundances are shown as the size of the circles.
    Unweighted PD takes the sum of the branch lengths connecting samples.
    Abundance-weighted measures take a weighted sum of branch lengths where weight is determined in some way by the abundance of the taxa on either side of the branch: if we give edges width according to their weight, the abundance-weighted measure can be thought of as the sum of the areas of the edges.
}
\forarxiv{
\begin{wrapfigure}[23]{r}{2.5in}
  % \capstart
  \vspace{-27pt}
  \begin{center}
    \includegraphics[width=2.5in]{figures/pd.pdf}
  \end{center}
  \vspace{-10pt}
  \caption{\pdLegend}
  \label{fig:pd}
\end{wrapfigure}
}

@@

\subsection{Diversity estimates using phylogenetics}

Because 16S surveys are inherently complex and noisy data, summary statistics are often used; summaries of the diversity of a single sample are often called \emph{alpha diversity}.
For the most part, this literature adapts methods from the classical ecological literature by substituting OTUs for taxonomic groups.
%The most common diversity indices are Simpson and Shannon.
% \citep{hill1973diversity} Diversity and evenness: a unifying notation and its consequences
However, phylogenetic diversity metrics are also used and here we will focus on their applications and methods.

Despite the enthusiasm with which microbial ecologists have accepted between-community comparison using phylogenetics (next section), phylogenetic alpha diversity seems under-developed.
Phylogenetic diversity (PD) measures use the structure and branch lengths of a phylogenetic tree to measure the diversity of a sample (Fig.~\ref{fig:pd}).
The original (unweighted) PD of a set of taxa is simply the total of the lengths of the branches connecting taxa in the set \citet{faith1992conservation}.
It quantifies the ``amount of evolution'' contained in the evolutionary history of those taxa.
Whereas just about every 16S survey investigation involves an OTU-based alpha diversity estimate, only a few involve PD: unweighted PD has been applied to some 16S survey data \citep{lozupone2007global,costello2009bacterial} and to metagenomic reads in a set of marker genes \citep{kembel2011phylogenetic}.

\newcommand{\unifracLegend}{
    The UniFrac divergence measure \citep[figure adapted from][]{LozuponeKnightUniFrac05}.
    When samples are interspersed across the tree (left tree), they have a smaller \textit{fraction} of branch lengths that can be \textit{uniquely} assigned to one sample or another compared to when they are separate (right tree).
}
\forarxiv{
\begin{wrapfigure}[23]{r}{2.5in}
  % \capstart
  \vspace{-16pt}
  \begin{center}
    \includegraphics[width=2.5in]{figures/unifrac.pdf}
  \end{center}
  \vspace{-10pt}
  \caption{\unifracLegend}
  \label{fig:unifrac}
\end{wrapfigure}
}

Although abundance weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are among the most common, abundance weighted phylogenetic diversity measures are not used in human microbiome studies.
Abundance-weighted measures take a sum of branch lengths weighted by abundance, such that branches that connect abundant taxa get a higher weight than ones that do not.
Thus rare taxa and artifactual sequences are down-weighted compared to abundant taxa.
Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
\citet{mccoy2013abundance} have recently shown that partially abundance weighted diversity measures do a good job of distinguishing between dysbiotic and "normal" states of the human microbiome; in particular that they do a better job than the commonly-used OTU-based measures.
\citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths.

@@

@@

Alignment methods are primarily focused on developing automated methods to extend a relatively small hand-curated ``seed alignment'' with additional sequences; several tools have been created with exactly this application for 16S in mind \citep{desantis2006nast,caporaso2010pynast,pruesse2012sina}.
The community also uses profile hidden Markov models \citep{eddy1998profile} and CM models \citep{nawrocki2009infernal,nawrocki2009structural} to achieve the same result.

The large data sets associated with human microbiome analysis require highly efficient algorithms for \emph{de novo} tree inference.
Historically this has meant relaxed neighbor joining \citep{evans2006relaxed}, but more recently FastTree 2 \citep{price2010fasttree} has emerged as the \textit{de facto} standard.
People do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has ported in the clearcut program \citep{sheneman2006clearcut}, or QIIME \citep{caporaso2010qiime}, which wraps clearcut and FastTree.
% AXIOME streamlines and manages analysis of small subunit (SSU) rRNA marker data in QIIME and mothur \citep{lynch2013axiome}.
% Bayesian phylogenetic inference is absent.

@@

Considerable effort goes to the creation of large curated alignments and phylogenetic trees on 16S.
There are two major projects to do so: one is the SILVA database \citep{pruesse2007silva,quast2013silva}, and the other is the GreenGenes database \citep{desantis2006greengenes,mcdonald2011improved}.
Because of the high rate of insertion and deletion of nucleotides in 16S, these alignments have a high percentage of gap.
Taking the length of 16S to be 1543 nucleotides,
% SILVA
% 1543 / 42284
% .96353230536373096207
the 479,726 sequence SILVA reference alignment version 115 is over 96\% gap,
% GreenGenes
% 1543 / 7682
% .79914084873730799272
while the 1,262,986 sequence GreenGenes 13\_5 alignment has is almost 80\% gap.
The SILVA-associated 'all-species living tree' project \citep{yarza2008all} started with a tree inferred by maximum likelihood and has been continually updated  by inserting sequences via parsimony.
The GreenGenes tree is updated by running FastTree from scratch for every release.
There appears to be a commonly held belief that FastTree in particular works well even with such gappy alignments \citep[e.g.][]{sharpton2011phylotu}.

@@

It appears that neutral models involving phylogenetics could be more fully developed.
Methods explicitly invoking trait evolution are notably absent, with the recent exception of \citep{langille2013predictive}.
The results of this simple method are reasonable, but would a collection of gene trees reconciled with a species tree allow for better prediction?
Perhaps improved methods, say involving whole-genome evolutionary modeling or models of metabolic network evolution, could shed light on the problem.
Here again the ``tree of life'' problem can be formulated in a practical light: what representation allows for the best prediction of features of underlying genomes?
How to formulate a useful notion of independent contrasts \citep{felsenstein1985phylogenies} on such an object?
It is quite possible that inference using a more complete representation would not be able to overcome the inherent noise of the data, but further exploration seems warranted as simple methods give reasonable results.

@@

\section{Acknowledgements}
I thank Olivier Gascuel for the opportunity to present on this subject during the 2013 LIRMM Mathematical and Computational Evolutionary Biology workshop and for organizing a corresponding special section of \textit{Systematic Biology}.
I am grateful to Aaron Darling, David Fredricks, Noah Hoffman, Steven Kembel, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, thank Bastien Boussau, Noah Hoffman, Christopher Small and Bj\"orn Winckler for providing feedback on the manuscript, and thank the NIH (R01 HG005966-01) and NSF (1223057) for financial support.

\notforarxiv{
\newpage
\section{Figure Legends}

\noindent
Figure \ref{fig:pd}: \pdLegend

\noindent
Figure \ref{fig:unifrac}: \unifracLegend

\noindent
Figure \ref{fig:dirtpiles}: \dirtpilesLegend
